---
title: "Jordan algebras in R"
author: "Robin K. S. Hankin"
output:
  pdf_document: default
  html_document:
    df_print: paged
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteIndexEntry{involutions}
  %\usepackage[utf8]{inputenc}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library("jordan")
```

# Jordan algebras

A _Jordan algebra_ is a nonassociative algebra over the reals with
a bilinear multiplication that satisfies the following identities:

$$xy=yx$$

$$(xy)(xx) = x(y(xx))$$

(the second identityis known as the Jordan identity).  In literature
one usually indicates multiplication by juxtaposition but one
sometimes sees $x\circ y$.  Package idiom is to use an asterisk, as in
`x*y`.  There are five types of Jordan algebras:

* Real symmetric matrices, class `real_symmetric_matrix`,
    abbreviated in the package to `rsm`
* Complex Hermitian matrices, class `complex_herm_matrix`,
    abbreviated to `chm`
* Quaternionic Hermitian matrices, class
    `quaternion_herm_matrix`, abbreviated to `qhm`
* Albert algebras, the space of $3\times 3$
    Hermitian octonionic matrices, class `albert`
* Spin factors, class `spin`

(of course, the first two are special cases of the next).  The
`jordan` package provides functionality to manipulate jordan objects
using natural R idiom.  Objects of all these classes are stored in
dataframe (technically, a matrix) form with columns being elements of
the jordan algebra.  The first four classes are matrix-based in the
sense that the algebraic objects are symmetric or Hermitian matrices
(the S4 class is `jordan_matrix`).  The fifth class, spin factors, is
not matrix based.

## Matrix-based Jordan algebras

These all behave in the same way from a package idiom perspective.
Consider:

```{r}
a <- rqhm()
a
```




```{r}
set.seed(0)
rspin()
```

This illustrates many features of the package. 